---
title: "Introduction to Census data"
format: html
---

::: summary
## Questions

-   How do we access census data?

## objectives

-   Combine census data with iNaturalist data
:::

Load packages

```{r}
#| warning: false
#| message: false

library(tidycensus)
library(dplyr)
library(mapview)
library(readr)
library(dplyr)
library(sf)
library(ggplot2)
```

## U.S. Census data

We are using tidycensus to access U.S. Census data.

## iNaturalist data

Let's get observations for the Great Blue Heron.

```{r}
#| cache: true
#| warning: false
#| message: false
inat <- read_csv('data/cleaned/cnc-los-angeles-observations.csv')
```

```{r}
inat_obs <- inat %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove=FALSE) %>% 
  filter(common_name == 'Great Blue Heron') 


```

```{r}
mapview(inat_obs)
```

## Median income by census tract

Let's get the median income for every census track in Los Angeles County from the 2020 American Community Survey.

```{r}
la_income <- get_acs(
  geography = "tract", 
  variables = "B19013_001",
  state = "CA", 
  county="Los Angeles",
  year = 2020,
  geometry = TRUE
)

```

The U.S. Census provides geographic data using the NAD 1983 geographic coordinate system.

```{r}
st_crs(la_income)
```

We want to convert the CRS of he census data to match the CRS of the iNaturalist data.

```{r}
la_income <- st_transform(la_income, crs = 4326)
```

compare the CRS

```{r}
st_crs(la_income) == st_crs(inat_obs)

```

Use ggplot2 to create a static map.

```{r}
ggplot() +
  geom_sf(aes(fill = estimate), data=la_income)   

```

Add iNaturalist data.

```{r}
ggplot() +
  geom_sf(aes(fill = estimate), data=la_income) +
  geom_sf(data=inat_obs)

```

Use mapview to create an interactive map.

```{r}
mapview(la_income, zcol = "estimate")

```

add iNaturalist dadta

```{r}
mapview(la_income, zcol = "estimate") +
  mapview(inat_obs)

```

## Population by census tract

Get population by census tract for Los Angeles County

```{r}
la_pop <- get_decennial(
  geography = "tract",
  variables = c(population="P1_001N"),
  state = "CA",
  county = "Los Angeles",
  year = 2020,
  geometry = TRUE
)
```

We want to convert the CRS of he census data to match the CRS of the iNaturalist data.

```{r}
la_pop <- st_transform(la_pop, crs = 4326)
```

Create static map

```{r}
ggplot() +
  geom_sf(aes(fill = value), data=la_pop)  +
  geom_sf(data = inat_obs)
```

Create interactive map

```{r}
mapview(la_pop, zcol="value") +
  mapview(inat_obs)
```

## Ethnicity by census tract

Get the percentage for each ethnicity for each census tract.

```{r}
la_ethnicity <- get_decennial(
  geography = "tract",
  state = "CA",
  county = "Los Angeles",
  variables = c(
    Hispanic = "P2_002N",
    White = "P2_005N",
    Black = "P2_006N",
    Native = "P2_007N",
    Asian = "P2_008N"
  ),
  summary_var = "P2_001N",
  year = 2020,
  geometry = TRUE
) %>%
  mutate(percent = 100 * (value / summary_value)) 
```

We want to convert the CRS of the census data to match the CRS of the iNaturalist data.

```{r}
la_ethnicity <- st_transform(la_ethnicity, crs = 4326)
```

Get the Hispanic data.

```{r}
la_hispanic <- filter(la_ethnicity,  variable == "Hispanic")
```

Create static map

```{r}
ggplot() +
  geom_sf(aes(fill = percent), data=la_hispanic) +
  geom_sf(data = inat_obs)
```

Create interactive map

```{r}
mapview(la_hispanic, zcol='percent') +
  mapview(inat_obs)

```
