## Creating Los Angeles geospatial files

Sometimes we have geospatial files that cover California or U.S., but we only want that data for LA County.

## State parks in LA area

California State Parks https://www.parks.ca.gov/?page_id=29682

California County Boundaries and Identifiers https://gis.data.ca.gov/datasets/California::california-county-boundaries-and-identifiers/explore

Read files

```{r}
parks_state <- read_sf('data/raw/Park_Boundaries/ParkBoundaries.shp')

ca_counties <- read_sf('data/raw/California_County_Boundaries_and_Identifiers_Blue_Version_view_2716745428963682036/California_County_Boundaries_and_Identifiers.shp')
```

get counties

```{r}
counties <- c('Los Angeles County', 
                       'Ventura County',
                       'San Bernardino County', 
                       'Orange County', 
                       'Riverside County', 
                       'Kern County', 
                       'Santa Barbara County', 
                       'San Luis Obispo County')

 socal_counties <-  ca_counties %>%
   filter( COUNTY %in% counties)
 
 la_county <-  ca_counties %>%
   filter( COUNTY == 'Los Angeles County')

mapview(socal_counties)
```

Compare CRS

```{r}
st_crs(parks_state) 
```

```{r}
st_crs(la_county)

```

Set and check CRS.

```{r}
la_county <- st_transform(la_county, crs = st_crs(parks_state) )
st_crs(parks_state) == st_crs(la_county)
```

Find CA parks within LA county

```{r}

la_parks <- parks_state %>% 
  st_intersection(la_county)

 mapview(la_county) +
 mapview(la_parks) 
```

```{r}



la_parks_2 <- parks_state[st_intersects(parks_state, la_county ) %>% lengths > 0,]

mapview(la_county) + mapview(la_parks_2)
```

create file

```{r}
st_write(la_parks_2, 'data/raw/ca_state_parks_la_county.geojson')
```

## National parks in LA area

National Park Boundaries https://catalog.data.gov/dataset/national-park-boundaries

Read files

```{r}
parks_nps <- read_sf('data/raw/nps_boundary/nps_boundary.shp')

ca_counties <- read_sf('data/raw/California_County_Boundaries_and_Identifiers_Blue_Version_view_2716745428963682036/California_County_Boundaries_and_Identifiers.shp')


```

get counties

```{r}
la_county <-  ca_counties %>%
   filter( COUNTY == 'Los Angeles County')
```

get CRS

```{r}
st_crs(parks_nps)
```

```{r}
st_crs(la_county)
```

set and check crs

```{r}
parks_nps <- st_transform(parks_nps, crs = st_crs(la_county) )
st_crs(parks_nps) == st_crs(la_county)
```

get national parks within LA county

```{r}

la_parks <- parks_nps[st_intersects(parks_nps, la_county ) %>% lengths > 0,]

mapview(la_county) + mapview(la_parks)
```

write file

```{r}
st_write(la_parks, 'data/raw/nps_la_county.geojson')
```
