---
title: "Creating charts"
format: html
include-in-header: "components/add-answer-button.html"
---

::: summary
## Questions

-   How do we create charts using R?

## Objectives

-   Create bar and line charts using ggplot2.
-   Customize the appearance of the charts.
-   Save charts as image files.
:::

```{r, load_packages}
#| message: false

library(readr) # read and write tabular data
library(dplyr) # manipulate data
library(ggplot2) # create data visualizations
library(lubridate) # manipulate dates
```

## Creating charts

**`ggplot2`** is a powerful package that allows you to create complex plots from tabular data (data in a table format with rows and columns). In the previous lesson we used it for maps, in this lesson we will use it for charts.

::: callout-note
Some people involved with visualizing data say charts, graphs, and plots are different things. This workshop will use charts, graphs, and plots interchangeably.
:::

Main steps:

1.  Load iNaturalist data
2.  Use `filter()`, `select()`, `mutate()`, and `count()` to get the rows and columns we want
3.  Pass the data to `ggplot2()`. Set which columns in a data frame will be used for chart elements such as x and y axis
4.  Create chart

## iNaturalist data

First, read data from the iNaturalist observation file.

```{r, load_inaturalist_data}
#| warning: false
#| message: true
inat_data <- read_csv('data/cleaned/cnc-los-angeles-observations.csv')

```

Let's add a year column to iNaturalist data.

```{r, add_year_column}
inat_year <- inat_data %>% 
  mutate(year = year(observed_on)) 
```

## Bar chart

Let's create a bar chart that shows the number of observations per year.

First, pass the data frame `inat_year` to `ggplot()` function as the first argument.

```{r, pass_data_to_ggplot}
ggplot(data = inat_year)
```

We need to tell **ggplot** how to use the data to create a chart. We can specify the mapping of data frame fields to elements in the chart, such as x/y coordinates, size, or shape, by using the `aes()` function. `aes` is short for aesthetic mappings.

For bar charts, we need to tell what field to use for the x axis. We want to create a plot with `year` on the x axis so we set the `aes()` function `x` argument to `year`. **ggplot** will count the number of rows for each year, and use the number of rows for the y axis.

```{r, pass_aes_to_ggplot}
ggplot(data = inat_year, mapping = aes(x = year))
```

Next we need to specify how we want the data to be displayed. We do this using `geom_` functions, which specify the type of geometry we want, such as points, lines, or bars. We use `geom_bar()` to create a vertical bar plot.

We can add a `geom_bar()` layer to our plot by using the `+` sign. We indent onto a new line to make it easier to read, and we have to end the first line with the `+` sign.

```{r, set_bar_type}
ggplot(data = inat_year,  mapping = aes(x = year)) +
  geom_bar()    

```

If we want horizontal bar chart, set the `aes()` function `y` argument to `year`.

```{r, horizontal_bar_chart}
ggplot(data = inat_year, mapping = aes(y = year)) +
  geom_bar()
```

## Line chart

Create a line chart that shows the number of observations per year.

For line charts, we need to specify which columns to use for the x and y axis.

We need to create a data frame that has a year column, and a column that counts the observations per year. Use `mutate()` to add year column. Use `count()` to add column that has the count per year.

```{r, create_dataframe_with_year_count}
inat_year_count <- inat_data %>% 
  mutate(year = year(observed_on)) %>%
  count(year, name='count')  

inat_year_count
```

For `aes`, use `year` on the `x` axis and `count` on the `y` axis. Use `geom_line()` to create a line chart.

```{r, create_line_chart}
ggplot(data = inat_year_count, 
       mapping = aes(x = year, y=count)) +
  geom_line()
```

## More bar charts

In the previous example we created a data frame `inat_year_count` that has observations per year. To create bar chart when the data has values for x axis and y axis, use `geom_col()`

We want `year` on the `x` axis, and `obs_count` on the y axis.

```{r, create_column_chart}
ggplot(data = inat_year_count, 
       mapping = aes(x = year, y = count)) +
  geom_col()
```

:::: exercise
## Exercise 1

Create a chart with all your observations for each year

-   Use `mutate()` to add year column

-   Use `filter()` to select observations you made

-   Save the data frame to `my_obs_by_year` object

-   Use `ggplot()` to set the data and aesthetics.

-   Choose which type of chart you want: bar or line.

::: answer
```{r, exercise_create_bar_chart}

my_obs_by_year  <- inat_data %>% 
  mutate(year = year(observed_on))  %>% 
  filter(user_login == 'natureinla')

  
ggplot(data = my_obs_by_year, mapping = aes(x = year)) +
  geom_bar()
```
:::
::::

## Changing colors

Use `colors()` to get a list of the 657 colors in R.

```{r, get_all_colors}
colors()
```

We can change the color of the bars using `fill`, `color`, and the name of the color. `fill` is the inside of the bar, `color` is the border of the bar.

```{r, set_bar_color}
ggplot(data = inat_year, mapping = aes(x = year)) +
  geom_bar(fill='aquamarine', color='black')
```

We can also use 6 digit hex color. You can use online tools to get [hex colors](https://html-color.codes).

```{r, set_bar_color_using_hex_code}
ggplot(data = inat_year,  mapping = aes(x = year)) +
  geom_bar(fill='#75cd5e')
```

## Color scales

In the previous bar chart, we made all the bars the same green color. If we want the chart colors to vary based on the data values, we need to put that information inside `aes()`, such as `aes(fill=<field_name>)`.

Let's update the previous bar chart to give each year a different color

```{r fill_warining_numerical_value}
ggplot(data = inat_year, 
       mapping = aes(x = year, fill = year)) +
  geom_bar()  
```

That didn't work. There is a warning: "Did you forget to specify a \`group\` aesthetic or to convert a numerical variable into a factor?"

::: callout-note
Factor is a R data structure used to represent categorical data, which only accepts a finite number of pre-defined values. To create a factor use `factor()`. The distinct values in a factor are called levels.

```{r creating_factor}
factor(c(2020, 2020, 2021, 2022, 2022, 2022))

factor(c('apple', 'grape', 'apple', 'pear'))
```
:::

Let's convert year into a factor using `factor(year)`.

```{r create_chart_with_fill_color_year}
ggplot(data = inat_year, 
       mapping = aes(x = year, fill= factor(year))) +
  geom_bar()  
```

**ggplot2** assigns each year a different color. **ggplot2** has color scale function that produces a predefined list of colors. The default color scale isn’t always ideal: it isn’t friendly to viewers with colorblindness and it doesn’t translate well to grayscale. However, **`ggplot2`** comes with quite a few other color scales, including the `viridis` scales, which are designed to be colorblind and grayscale friendly. We can change color scales by adding `scale_color_`  and `scale_fill_` functions to our plots. Since we are dealing with bar charts, we want `scale_fill_.`

::: callout-note
In mathematics and statistics, values can be discrete or continuous. Discrete value have distinct values such as whole numbers (1, 2, 3) or categories (states: California, Oregon). Continuous values represent an infinite number of possible values such as fractions and decimals (1.1, 2-1/2, 2-3/4, 3.9).
:::

Color scale functions can be discrete or continuous. Let's add viridis scale to the previous bar chart. We want to use `scale_fill_viridis_d()` since year has discrete values.

```{r create_chart_with_fill_viridis_year}
ggplot(data = inat_year,  
       mapping = aes(x = year, fill= factor(year))) +
  geom_bar()  +
  scale_fill_viridis_d()
```

 If you try to use viridis continuous funtion, you will get an error: "Discrete values supplied to continuous scale."

```{r create_chart_with_fill_viridis_year_error}
#| eval: false 
#| 
ggplot(data = inat_year, 
       mapping = aes(x = year, fill= factor(year))) +
  geom_bar()  +
  scale_fill_viridis_c()
```

For more information about color scales, visit [Introverse color scales](https://sjspielman.github.io/introverse/articles/color_fill_scales.html) page.

## Changing axis breaks

**ggplot2** will automatically break the axis into multiple chunks and add labels for the values. We can use `n.breaks` argument for `scale_x_continuous()` and `scale_y_continuous()` to adjust the number of breaks along the axis. **ggplot2** may choose a slightly different number to ensure nice break labels.

Let's add more breaks along the year x axis and count y axis.

```{r create_chart_with_custom_breaks}
ggplot(data = inat_year,  
       mapping = aes(x = year, fill= factor(year))) +
  geom_bar()  +
  scale_fill_viridis_d() +
  scale_x_continuous(n.breaks=8) +
  scale_y_continuous(n.breaks=7)
```

Now we have labels for every year and every 5000 observations.

## Changing themes

Let's assign a plot to an object so can refer to it later.

```{r, assign_plot_to_object}
myplot <- ggplot(data = inat_year,  
                 mapping = aes(x = year)) +
  geom_bar(fill='#75cd5e') +
  scale_x_continuous(n.breaks = 8) 
 

  myplot
```

**ggplot** has built in themes that affects the overall appearance of a plot. To see a list of available themes in ggplot, visit [ggplot2 themes](https://ggplot2.tidyverse.org/reference/index.html#themes)

Let's add the black-and-white theme `theme_bw()` to our plot:

```{r set_theme, warning=FALSE,}
myplot + 
  theme_bw()
```

To individually change parts of a plot, we can use the `theme()` function, which can take many different arguments to change things about the text, grid lines, background color, and more.

Let's rotate the x axis labels by 90 degrees so the text is vertical.

```{r rotate_axis_label}
myplot + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))

```

Let's try changing the size of the text on our axis titles. We can do this by specifying that the `axis.title` should be an `element_text()` with `size` set to 14.

```{r set_text_size, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14))
```

Another change we might want to make is to remove the vertical grid lines. To do this, we will change the `panel.grid.major.x` and `panel.grid.major.x` to an `element_blank()`.

```{r remove_vertical_grid_lines, warning=FALSE}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

::: callout-tip
Because there are so many possible arguments to the `theme()` function, it can sometimes be hard to find the right one. Here are some tips for figuring out how to modify a plot element:

-   type out `theme()`, put your cursor between the parentheses, and hit <kbd>Tab</kbd> to bring up a list of arguments
    -   you can scroll through the arguments, or start typing, which will shorten the list of potential matches
-   like many things in the `tidyverse`, similar argument start with similar names
    -   there are `axis`, `legend`, `panel`, `plot`, and `strip` arguments
-   arguments have hierarchy
    -   `text` controls all text in the whole plot
    -   `axis.title` controls the text for the axis titles
    -   `axis.title.x` controls the text for the x axis title
:::

## Changing labels

We can customize the axis labels and add a chart title using `labs()`.

```{r changing_labels, warning=FALSE}
myplot +
  labs(title = "CNC Los Angeles",
       subtitle="Observations per year",
       x = "Year",
       y = "Observations")
```

We can also change the title in the legend using `labs()`

```{r}
ggplot(data = inat_year,  
       mapping = aes(x = year, fill= factor(year))) +
  geom_bar() +
  labs(fill = "Year")
```

:::: exercise
## Exercise 2: Customizing a plot

Take the `my_obs_by_year` data frame from exercise 1 and create a plot. Customize the appearance or the chart. Here are some ideas.

-   add a title

-   choose the one of the built in themes

-   change the axis titles

-   change the fill colors of the bars

::: answer
```{r, exercise_change_chart_appearance}
my_chart <- ggplot(data = my_obs_by_year, 
                   mapping = aes(x = year)) +
  geom_bar()

my_chart +
  theme_classic() +
  labs(title = "CNC Los Angeles",  x = "Year",  y = "Observations") 
```
:::
::::

## Saving plots

Once we are happy with ourplot, we can save the plot.

Assign the plot to an object. Then run `ggsave()` to save our plot. The first argument we give is the path to the file we want to save, including the correct file extension. You can save as jpb, pdf, tiff, png. Next, we tell it the name of the plot object we want to save. We can also specify things like the width and height of the plot in inches.

```{r, save_plot}
#| eval: false

# create plot
finalplot <- myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.title = element_text(face = "bold", size = 20)) +
  labs(title = "CNC Los Angeles",
       subtitle="Observations per year",
       x = "Year",
       y = "Observations")

# save plot
ggsave(filename = 'results/observations_per_year.jpg', plot = finalplot, 
       height = 6, width = 8)
```
