---
title: "Setup"
format: html
---

This workshop use Github Codespaces to host the software, data, and code needed for the workshop. This way attendees won't need to install stuff on their own computer.

If you are interested in install all the software, data, and code on your computer, visit the [Setup - local page](setup-local.qmd)

## Functions

Functions are predefined bits of code that do a specific task. Arguments are values that we pass into a function. Function usually takes one or more arguments as input, does something to the values, and produces the ouput.

## R packages

R itself has many built-in functions, but we can access many more by installing and loading other **packages** of functions and data into R. We will use several R packages for the workshop.

To install these packages, use `install.packages()` function from R. We pass in the package names as arguments. The name of the packages must be in quotes.

```{r, install_packages}
#| eval: false
install.packages("readr")
```

R will connect to the internet and download packages from servers that have R packages. R will then install the packages on your computer. The console window will show you the progress of the installation process.

To save time, we have already installed all the packages we need for the workshop.

In order to use a package, use `library()` function from R to load the package. We pass in the name of the package as an argument. Do not use quotes around the package name when using `library()`.

```{r load_package}
#| eval: false
library(readr) 
```

## Workshop R packages

These are the R packages we will use in this workshop.

```{r load_all_libraries}
#| eval: false
library(readr) # read and write tabular data
library(dplyr) # manipulate data
library(ggplot2) # create data visualizations
library(sf) # handle geospatial data
library(lubridate) # manipulate dates
library(mapview) # create interactive maps

```
