---
title: "Understanding data"
format: html
include-in-header: "components/add-answer-button.html"
---

::: summary
## Questions

-   How does R store and represent data?

## Objectives

-   Understand data types and missing values
-   Learn about data structures vectors and data.frame
:::

We started the previous lessons with `read_csv()`. To better understand the data returned by `read_csv()`, we will learn how R represents and stores data.

```{r, loading_packages}
#| message: false

library(readr) # read and write tabular data
library(dplyr) # manipulate data
```

```{r, assign_read_csv_to_object}
#| message: false
#| warning: false

inat_data <- read_csv('data/cleaned/cnc-los-angeles-observations.csv')

```

Let's look at the data information returned by `glimpse()`.

```{r, call_glimpse}
glimpse(inat_data)
```

## Data Types

`glimpse()` shows `<dbl> <date> <chr> <lgl>`. Those data types.

In computer programming, **data type** is a way to group data values. Every value has a data type. A value's data type will determine what the programming language can do with the value. For instance, in R we can add numbers but we can't add strings.

There are 4 main types in R: **numeric**, **integer**, **logical**, and **character**.

-   **numeric** are numbers that contain a decimal (e.g. 1.2, 10.5). By default, R also treats whole numbers as decimals (e.g. 1, 10).

-   **integer** are whole numbers that do not have a decimal point. (e.g. 1L, 10L). In R, the `L` suffix forces the number to be an integer, since by default R uses decimal numbers.

-   **logical** have values of `TRUE` or `FALSE`.

-   **character** are strings of characters (e.g. "abc", 'dog'). The strings must be surrounded by quotes, either single quotes or double quotes.

## Vectors

A **data structure** is a way to organize and store a collection of values.

A **vector** is data structure in R that has series of values. All the value in the vector must be the same data type.

To create a vector we use the `c()` combine function, and pass in the values as arguments. We can use `class()` function to find the type or class of any object.

numeric vector

```{r numeric_vector}
numbers <- c(1, 2, 5)
numbers

class(numbers)
```

character vector

```{r character_vector}
characters <- c("apple", 'pear', "grape")
characters

class(characters)
```

logical vector

```{r logical_vector}
logicals <- c(TRUE, FALSE, TRUE)
logicals

class(logicals)
```

If you try to put values of different data types into a vector, all the values are converted to the same data type. In the following example, everything is converted to character type.

```{r vector_conversion}
mixed <- c(1, "apple", TRUE)
mixed

class(mixed)
```

## Missing data

R represents missing data as `NA`, without quotes. `NA` stands for not applicable.

`NA` is allowed in vectors of any data type.

```{r c-vector-na}
numbers <- c(1, 2, NA)
numbers

class(numbers)
```

When you pass number vectors with `NA` to a math function like `min()` (returns minimal value in a vector), the function returns `NA`.

```{r}
min(numbers)
```

Many math functions have an argument `na.rm` to remove `NA` values for its calculations.

```{r}
min(numbers,  na.rm = TRUE)
```

## data.frame

When we call `class()` with `inat_data`, it returns "tbl" and "data.frame".

```{r}
class(inat_data)
```

**data.frame** is a data structure from R that is used to represent tabular data with rows and columns. **tbl** aka tibble is an updated version of data.frame from the **tibble** package. **readr** uses tibbles instead of data.frames.

When `readr` loads a CSV, it tries to figure the data type for each column. Each column in a data.frame or tibble is a vector. Because each column is a vector, all the values in a column must be of the same data type.

For `inat_data`, `id` is **numeric**, `captive_cultivated` is **logical**, `user_login` is **character**. Multiple columns have `NA` such as `license` and `sound_url`.

`readr` has a special data type `date` that is used to represent dates. Column `observed_on` is **date**. Other columns are such as `time_observed_at` are treated as **character** because those strings have extra information that `readr` does not recognize as a **date**.
